# Tidy Tuesday Data

The R for Data Science Community releases data that can be used to practice data science skills.

If you do decide to use any of these data set, we kindly ask to fill the R4DS Survey to indicate that the data set are being used in class.

## Taylor Swift Data

The `taylor` package provides the `taylor_all_songs` and `taylor_album_songs` data sets that contain information on all of Taylor Swift's songs. Each data set contains basic information of songs (song release date, album releaser, ...) and Spotify API level data (danceability score, energy score, ...). You can install the `taylor` package with the following code:

```{r}
#| echo: true
#| eval: false
install.packages("taylor")

```

Access the data with:

```{r}
#| echo: true
#| eval: false

taylor::taylor_album_songs

## OR

library(taylor)
taylor::taylor_album_songs
```

The help documentation can be found here: <https://taylor.wjakethompson.com/>

### Data

One hundred randomly selected songs are displayed below:

```{r}
#| echo: false

taylor::taylor_album_songs |> 
  dplyr::slice_sample(n = 100) |> 
  DT::datatable(options = list(dom = "tp", pageLength = 5))

```

## Anime

The `anime` data set from MyAnimeList contains information on rankings and popularity scores of different anime episodes. More information can be found [here.](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-04-23) The data set contains 77,911 observation and 28 variables. **Warning: This data set will take over 30 MB to load up in R.**

```{r}
#| message: false
anime <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-04-23/tidy_anime.csv")
```

### Data

Here is a sample of the data set, with the `synopsis` variable removed to display the table better.

```{r}
#| echo: false
anime |> 
  dplyr::slice_sample(n=100) |> 
  dplyr::select(-synopsis) |> 
  DT::datatable(options = list(dom = "tp", pageLength = 3))

```

## Starbucks

```{r}
#| message: false

starbucks <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-12-21/starbucks.csv')

```

### Data

A random sample of 100 products are shown.

```{r}
starbucks |> 
  dplyr::slice_sample(n=100) |> 
  DT::datatable(options = list(dom = "tp", pageLength = 5))
```
